\chapter{TODO}
\section*{12 - Ottobre}
\section{Component}
Element of implementing a coherent set of features
...

Architectural design issues
\begin{itemize}
    \item Non-functional product characteristics: security, reliability, availability...
    These aspects are as important as functional properties.
    \item Product lifetime: 
    \textit{microservices}, for instance, easily allow scalability increasing the lifetime of our product.
    \item Software compatibility:
    Usually there may be legacy modules in the system,
    thus compatibility may be crucial.
    \item Number of users:
    Releasing software on the internet truly complicates the prediction of the number of users for a product 
    \item Software reuse: 
\end{itemize}

\section{Non-functional quality attributes}
\begin{enumerate}
    \item Responsiveness
    \item Reliability\footnote{Do features behave as expected?}
    \item Availability
    \item Security
    \item Usability
    \item Maintainability
    \item Resilience
\end{enumerate}
This typically aren't features in the mid-development prototypes,
they usually refer to the final product.
Implementing these in the prototypes would increase too much the time taken to develop such prototypes.
\nl

System decomposition must be done in conjunction with choosing technologies for the system.
In this there is a mixup happening between implementation and design, however it is necessary.

\section{Client-Server architecture}
aka Model-View controller where usually the communication between client and server happens with HTTP along with JSON (/XML).
Client requests to a server are then muxed on many slave nodes which elaborate the requests.


\section{Servre}