\chapter{Deadlocks}

\begin{definition}
   [Deadlock]
   A deadlock occurs when a set of processes is waiting for resources held by other processes in the set, resulting in a circular wait where no process can proceed.
\end{definition}

{There are Four \textit{Necessary} Conditions (\textit{Coffman} Conditions) in order to have a deadlock:\ns
\begin{itemize}
	\item \textbf{Mutual Exclusion}: Resources cannot be shared.
	\item \textbf{Hold and Wait}: Processes holding resources can request new ones.
	\item \textbf{No Preemption}: Resources cannot be forcibly taken away.
	\item \textbf{Circular Wait}: A closed chain of processes exists, where each holds a resource needed by the next.
\end{itemize}
If any of the previous conditions is not met, a deadlock cannot occur.}

{There are three main approaches to handle deadlocks, which act at different points in the deadlock cycle, but in reality the last one is the only true and available solution; Prevention and Avoidance typically come at a way too high cost:\ns
\begin{itemize}
   \item \textbf{Prevention}: Ensures that at least one of the necessary conditions does not hold.
   \begin{itemize}
      \item Acquire all resources before starting.
      \item Preempt a process holding a needed resource.
   \end{itemize}
   This approach typically leads to \ul{resource underutilization and low throughput}.
   \item \textbf{Avoidance}: Checks dynamically if granting a resource would lead to a deadlock.
   \begin{itemize}
   	\item Banker's Algorithm (used in centralized systems).
	   \item Safe state evaluation in distributed environments.
   \end{itemize}
   Avoidance requires \textit{global state awareness}.
   \item \textbf{Detection and Recovery}: Identifies a deadlock state and then recovers from it.
   \begin{itemize}
      \item Once detected, deadlocks can be resolved by aborting one or more processes or preempting resources
      \begin{itemize}
         \item \textbf{Process Termination} - Abort one or more processes involved in the deadlock.
         \item \textbf{Resource Preemption} - Forcefully take a resource from one process to give it to another.\\
         It is complex to implement without causing inconsistencies. It requires the ability to roll back (restore) the state of the preempted process.
         \item \textbf{Rollback} - Roll back the state of one or more processes to a safe point before the deadlock occurred.
      \end{itemize}
   \end{itemize}
\end{itemize}}


\section{RAG - Resource Allocation Graph}
The Resource Allocation Graph is a directed graph that models the allocation of resources to processes. It is composed of two types of nodes, processes and resources, and two types of edges, request and assignment. 
A cycle in the graph indicates a deadlock.

\subsection{WFG - Wait-For Graph}
A Wait-For Graph is a directed graph, a simplified RAG, that models the wait-for relationship between processes. It is composed of processes as nodes and edges representing the wait-for relationship. A cycle in the graph indicates a deadlock.

\section{Detecting Deadlocks}

There are some challenges in detecting deadlocks in distributed systems, such as the lack of a global state, communication delays, and the need for coordination among nodes. More specifically, we risk:
\begin{itemize}
   \item False positives due to delays in the netowrk or inconsistent state information
   \item Overhead due to message passing
   \item Dynamic changes in resource allocation
\end{itemize}

Deadlock detection in distributed systems involves monitoring process and resource states across different nodes and identifying circular wait conditions globally.\\
{Correctness Conditions are Progress and Safety, which are, respectively:\ns
\begin{itemize}
	\item Progress: The algorithm must detect all deadlocks in finite time.
	\item Safety: No false deadlocks (called phantom or false deadlocks) should be reported.
\end{itemize}
}
\subsection{Centralized Deadlock Detection}
A central node gathers information from all other nodes and constructs a global wait-for graph.
It is very simple, but not scalable, and the node can become a single point of failure.
\subsection{Hierarchical Deadlock Detection}
The system is divided into regions, with each region responsible for local deadlock detection.  Regional coordinators communicate with higher-level nodes for global detection.

This approach provides better scalability than centralized detection, but, on the other hand, coordination between regions can be complex.
\subsection{Distributed Deadlock Detection}

Here all nodes participate in deadlock detection by sharing partial information about resources and processes, and each node detects deadlocks locally and communicates with others to detect global deadlocks.

There is no single point of failure, and the approach scales well with larger systems, but poses more complex message passing and increased overhead.

\subsection{System Model for Deadlock Detection - WFG}
The distributed system is composed of asynchronous processes that communicate via message passing.
A WFG models the state of the system.

\begin{itemize}
	\item Nodes:
	\begin{itemize}
      \item 
      Processes.
   \end{itemize}
	\item Edges:
	\begin{itemize}
      \item 
      Directed edges from P1 to P2 indicate
      that P1 is waiting for P2 to release a
      resource.
   \end{itemize}
	\item Deadlock occurs when there’s a cycle in the WFG
\end{itemize}
\labelitemize{
   Assumptions
}{
   \begin{itemize}
      \item Only reusable resources.
      \item Exclusive resource access.
      \item One copy of each resource.
      \item Two process states: Running (active) or Blocked (waiting for resources).
   \end{itemize}
}

The challenges in this system are \textbf{WFG maintenance}, which involves tracking resource dependencies across distributed nodes, and \textbf{cycle detection}, which involves searching cycles in the WFG. 
In particular, such search should find \textit{all} cycles in \textit{finite} time. Besides no false ---aka \textit{phantom}--- deadlocks should be reported.

{Having the Deadlocks detected, the system can then proceed to resolve them by \textbf{rollback}, \textbf{preemption}, or \textbf{termination}.\ns
\begin{itemize}
	\item Break the wait-for dependencies between deadlocked processes.
	\item Roll back or abort one or more processes to free up resources.
	\item Clean up the wait-for graph after resolution to avoid phantom deadlocks.
	\item \ul{Untimely and inappropriate cleaning of broken wait-for dependencies is the main reason why many deadlock detection algorithms reported in the literature are incorrect}
\end{itemize}}

\section{Deadlock Models}
\subsection{Single Resource Model}
In this model, each process requires a single resource to proceed. The deadlock detection algorithm is simple and efficient, but it is not very realistic.
Since the maximum out-degree of a node in a WFG for the single resource model can be 1, the presence of a cycle in the WFG shall indicate that there is a deadlock

\subsection{AND Model}
A process can request multiple resources simultaneously, and the request is granted only if all resources are available.
A cycle in the WFG ---guess what--- implies a deadlock.\\
In the AND model, if a cycle is detected in the WFG, it implies a deadlock but \textit{\textbf{not} vice versa}. That is, a process may not be a part of a cycle, it can still be deadlocked.

\subsection{OR Model}
In the OR model, a process can make a request for numerous resources simultaneously and the request is satisfied if any one of the requested resources is granted.\\
Presence of a cycle in the WFG of an OR model does \textbf{not} imply a deadlock in the OR model.

In the OR model, the presence of a \textbf{knot} indicates a deadlock.
% In a WFG,a vertex $v$ is in a knot if for all $u::u$ is reachable from $v:v$ is reachable from $u$.
A knot is a set of processes such that each process in the set is waiting for a resource held by another process in the set.\\
More formally, a knot $K$ is a non-empty subset of processes such that:
\begin{itemize}
      \item \textbf{Mutually reachable} - $\forall p,q \in K$ $p$ is reachable from $q$ and $q$ is reachable from $p$ ($K$ is \textit{strongly connected}).
      \item \textbf{No dead ends} - $\forall p \in K$, if $p$ is waiting for a process $r$, then $r \in K$. In other words, there are no outgoing edges from processes in $K$ to processes outside $K$ (which could potentially lead to resource acquisition and deadlock resolution), so $K$ is closed with respect to reachability towards processes that can provide resources
\end{itemize} 
% No paths originating from a knot shall have dead ends.

Note that, \ul{there can be a deadlocked process that is \textbf{not} a part of a knot}.\\
So, in an OR model, a blocked process $P$ is deadlocked if it is either in a knot or it can only reach processes on a knot.
\note{Remember that graph is directed, so \ul{reachability is \textbf{not} symmetric}. A process $p$ can reach process $q \in K$ if there is a directed path from $p$ to $q$, but the reverse may not necessarily be true, so the ``\textbf{No dead ends}'' condition would still hold}

\section{Deadlock Detection Algorithms}
\subsection{Path-Pushing Algorithm - Global WFG}
Distributed deadlocks are detected by
maintaining an explicit global WFG
\begin{itemize}
	\item to build a global WFG for each site of the
distributed system
	\item Nodes propagate local WFG to other nodes.
	\item After the local data structure of each site is
updated, WFG are updated
	\item Deadlocks are detected by identifying circular
dependencies.
\end{itemize}

\subsection{Edge-Chasing - Probing}
Here, a process sends probes to check if it is part of a deadlock.
Whenever a process that is executing receives a probe message, it simply discards this message and continues. Only blocked processes propagate probe messages along their outgoing edges\footnote{I suppose of a WFG}. If the probe returns to the initiating process, a deadlock is detected.

\begin{itemize}
	\item P1 sends a \textit{probe} to P2, which forwards it to P3.
	\item When the \textit{probe} returns to P1, a deadlock is detected
\end{itemize}

A probe is a message used in edge-chasing algorithms to \ul{trace the path of resource dependencies}.

\subsection{Diffusion Computation Algorithm - Query Propagation}
\begin{itemize}
	\item An algorithm for deadlock detection based on diffusion computation.
	\item When a process is blocked, it propagates queries to other processes, otherwise it drops queries
	\item Queries are discarded by a running process and are echoed back by blocked processes in the following way:
   \begin{itemize}
   	\item When a blocked process first receives a query message for a particular deadlock detection initiation, it does not send a reply message until it has received a reply message for every query it sent (to its successors in the WFG).
   	\item For all subsequent queries for this deadlock detection initiation, it immediately sends back a reply message.
   	\item The initiator of a deadlock detection detects a deadlock when it has received a reply for every query it has sent out.
   \end{itemize}
\end{itemize}

\subsection{Global state detection-based algorithms}

These algorithms exploit the following facts:
\begin{itemize}
	\item a consistent snapshot of a distributed system can be obtained without freezing the underlying computation
	\item a consistent snapshot may not represent the system state at any moment in time, but if a stable property holds in the system before the snapshot collection is initiated, this property will still hold in the snapshot.
\end{itemize}

Therefore, distributed deadlocks can be detected by taking a snapshot of the system and examining it for the condition of a deadlock.



\subsection{Deadlock Detection in Dynamic and Real-time Systems}

In distributed systems with dynamic resources (e.g., cloud systems), resource requests and releases are constantly changing. Deadlock detection algorithms must adapt to these changes and avoid outdated information.

This clearly poses consistent challenges such as false detections, the need for constant
updates, and handling frequent resource changes.


In real-time distributed systems, deadlocks must be detected and resolved quickly to meet timing constraints.\\
Detection algorithms must have low latency and minimal overhead

\section{Assignment}

Study and eventually prepare a few slides on
\begin{itemize}
   \item Misra-Chandy-Haas Algorithm for AND model
   \item Misra-Chandy-Haas Algorithm for OR model
\end{itemize}

In Misra-Chandy-Haas algorithm a process suspecting a deadlock sends a \textbf{probe} $\langle P_0,P_i,P_k\rangle$, where:
\begin{itemize}
   \item $P_0$ is the process that initiated the probe
   \item $P_i$ is the process that at the current iteration sent the probe to $P_k$
   \item $P_k$ is the process that $P_i$ is waiting for
\end{itemize}
When the probe returns to $P_0$, a deadlock is detected and confirmed.

The key difference between the AND and OR models is in the way the \ul{probe is propagated}. In the AND model, the probe is propagated along \textbf{all} outgoing edges, while in the OR model, the probe is propagated along \textbf{only one} outgoing edge.

\subsection{Misra-Chandy-Haas Algorithm - Detailed Explanation}

The Misra-Chandy-Haas algorithm is an edge-chasing, probe-based deadlock detection algorithm that works in distributed systems. It is designed to detect cycles in the Wait-For Graph (WFG) by propagating probe messages along dependency edges.

\subsubsection{Key Concepts}
\begin{itemize}
   \item \textbf{Probe Message}: A triplet $\langle P_0, P_i, P_k \rangle$ where:
   \begin{itemize}
      \item $P_0$ is the \textit{initiator} (the process suspecting it might be deadlocked)
      \item $P_i$ is the \textit{sender} (the process currently forwarding the probe)
      \item $P_k$ is the \textit{receiver} (the process that $P_i$ is waiting for)
   \end{itemize}
   \item \textbf{Probe Propagation}: \ul{Only blocked processes propagate probes; running processes discard them immediately}.
   \item \textbf{Deadlock Detection}: If a probe returns to its initiator $P_0$ (i.e., $P_0$ receives a probe $\langle P_0, P_j, P_0 \rangle$), a cycle is detected and a deadlock is confirmed.
\end{itemize}

\subsubsection{Algorithm for AND Model}
In the AND model, a blocked process must wait for \textit{all} resources it has requested. Therefore, when a blocked process $P_k$ receives a probe $\langle P_0, P_i, P_k \rangle$:
\begin{enumerate}
   \item If $P_k = P_0$, a deadlock is detected (the probe has completed a cycle).
   \item If $P_k$ is running, it discards the probe.
   \item If $P_k$ is blocked and waiting for processes $\{Q_1, Q_2, \ldots, Q_n\}$, it \ul{forwards the probe to \textbf{all} of them}:
   \[
   \langle P_0, P_k, Q_1 \rangle, \quad \langle P_0, P_k, Q_2 \rangle, \quad \ldots, \quad \langle P_0, P_k, Q_n \rangle
   \]
\end{enumerate}

This exhaustive propagation ensures that all dependency paths are explored, which is necessary because in the AND model, \textit{all} dependencies must be satisfied.

\subsubsection{Algorithm for OR Model}
In the OR model, a blocked process needs only \textit{one} of the requested resources to proceed. Therefore, when a blocked process $P_k$ receives a probe $\langle P_0, P_i, P_k \rangle$:
\begin{enumerate}
   \item If $P_k = P_0$, a potential deadlock is detected.
   \item If $P_k$ is running, it discards the probe.
   \item If $P_k$ is blocked and waiting for processes $\{Q_1, Q_2, \ldots, Q_n\}$, it \ul{forwards the probe to \textbf{only one} of them} (typically following a single outgoing edge):
   \[
   \langle P_0, P_k, Q_j \rangle \quad \text{for some } j \in \{1, \ldots, n\}
   \]
\end{enumerate}

This selective propagation reflects the semantics of the OR model: since any one resource can break the wait, exploring a single path is sufficient to detect a knot. However, detecting all deadlocked processes in the OR model is more complex, as a process can be deadlocked even if not part of a cycle (if all its paths lead to a knot).

\subsubsection{Example: AND Model}
Consider processes $P_1 \to P_2 \to P_3 \to P_1$ (a cycle). 
\begin{itemize}
   \item $P_1$ initiates a probe $\langle P_1, P_1, P_2 \rangle$.
   \item $P_2$ (blocked, waiting for $P_3$) forwards $\langle P_1, P_2, P_3 \rangle$.
   \item $P_3$ (blocked, waiting for $P_1$) forwards $\langle P_1, P_3, P_1 \rangle$.
   \item $P_1$ receives the probe with itself as the receiver $\Rightarrow$ deadlock detected.
\end{itemize}

\subsubsection{Example: OR Model}
Consider $P_1 \to P_2$, $P_1 \to P_3$, $P_2 \to P_3$, $P_3 \to P_1$ (a knot).
\begin{itemize}
   \item $P_1$ initiates $\langle P_1, P_1, P_2 \rangle$ (choosing one edge).
   \item $P_2$ forwards $\langle P_1, P_2, P_3 \rangle$.
   \item $P_3$ forwards $\langle P_1, P_3, P_1 \rangle$.
   \item $P_1$ receives the probe $\Rightarrow$ potential deadlock (knot) detected.
\end{itemize}

\subsubsection{Advantages and Limitations}
\begin{itemize}
   \item \textbf{Advantages}:
   \begin{itemize}
      \item Distributed and scalable (no central coordinator).
      \item Low message overhead compared to global WFG construction.
      \item Detects deadlocks efficiently by tracing dependency chains.
   \end{itemize}
   \item \textbf{Limitations}:
   \begin{itemize}
      \item In the OR model, detecting all deadlocked processes (including those not in a cycle but reachable only to a knot) is more complex.
      \item False positives (phantom deadlocks) can occur if the WFG changes during probe propagation.
      \item Requires careful synchronization to avoid race conditions.
   \end{itemize}
\end{itemize}